---
title: "Exploratory Data Analysis"
output: github_document
date: "2025-10-02"
---
```{r}
library(tidyverse)
library(p8105.datasets)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


Import Weather Data
```{r}
data("weather_df")


weather_df = 
  weather_df |> 
  mutate(month = floor_date(date, unit = "month")) 


#floor(date() --> this takes the date and makes it a month. Ex.Feb 21st becomes Feb 1st 
#Look into package lubridate --> all about how to work with dates
```

Make plots

```{r}

weather_df |> 
  ggplot(aes(x=prcp)) +
  geom_histogram()

#right off the bat we can see some extreme values.
```

First thing, check extreme values

```{r}
weather_df |> 
  filter(prcp >1000)
```


Look at data again

```{r}
weather_df |> 
  filter(tmax >= 20, tmax <= 30) |> 
  ggplot(aes( x = tmin, y = tmax, color = name)) +
  geom_point()

#check out why waterhole is looking different than Malakai and NYC? It is because of how they got their temp.
```

Group by name
```{r}
weather_df |> 
  group_by(name)
#gives three distinct groups

weather_df |> 
  group_by(name, month)
#gives 72 becuase 12 months and 3 groups 
```


Sumarize groups 

```{r}
weather_df |> 
  group_by(name) |> 
  summarize(
    n = n()
  )

#prints out tibble



weather_df |> 
  group_by(name, month) |> 
  summarize(
    n = n()
  )
#n tells us how many days are in each month here


#distinct instances
weather_df |> 
  group_by(month) |> 
  summarise(
    n = n_distinct(date)
  )
```


Special function to count directly

```{r}
weather_df |> 
  count(name) #short hand that says group by name and count by group
```


```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE),
    median_tmin = median(tmin, na.rm = TRUE),
    sd_prcp = sd(prcp, na.rm = TRUE)
  ) |> view()

```

This is still a dataframe

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE)
    ) |> 
  ggplot(aes(x = month, y = mean_tmax, color = name)) +
  geom_point() +
  geom_line()
  
  
```

Sometimes you need to format results more nicely

```{r}
weather_df |> 
  group_by(name, month) |> 
  summarise(
    mean_tmax = mean(tmax, na.rm = TRUE)
  ) |> 
  pivot_wider(
    names_from = name,
    values_from = mean_tmax
  ) |> 
  knitr::kable(digits = 2)
```


# Mutate with groups

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    mean_tmax = mean(tmax, na.rm = TRUE),
    center_tmax = tmax - mean_tmax
  ) |> 
  ggplot(aes(x = date, y = center_tmax, color = name)) +
  geom_point()

#we are standardizing --> shifting our means so they equal zero 
```

Look for cold days

```{r}
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = min_rank(desc(tmin))) |> 
  filter(temp_rank < 2)

#without filtering every day would be ranked by tmin
weather_df |> 
  group_by(name, month) |> 
  mutate(temp_rank = min_rank(tmin))
```

What about lags?

```{r}
weather_df |> 
  mutate(lagged_tmax = lag(tmax))
#current form is not taking into account that we have different locations, grouping fixes this
#by default lag() is 1. what is the previous or next value you are interested in 


weather_df |> 
  group_by(name) |> 
  mutate(lagged_tmax = lag(tmax))
```


Use the variables you create

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax)
  ) |> 
  summarise(
    sd_tmax_change = sd(temp_change, na.rm = TRUE),
    tmax_change_max = max(temp_change, na.rm = TRUE)
  )
```


Figure out what day it happened

```{r}
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax),
    change_rank = min_rank(desc(temp_change))
    )|> 
  filter(change_rank < 2)


#Slightly slimmer version 
weather_df |> 
  group_by(name) |> 
  mutate(
    temp_change = tmax - lag(tmax)
    )|> 
  filter(min_rank(desc(temp_change)) < 2)

```

Let's revist the pulse dataset

```{r}
pulse_df =
  haven::read_sas("data/public_pulse_data.sas7bdat") |>  
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  mutate(visit = replace(visit, visit == "bl", "00m"),
         visit = fct_inorder(visit)
        )  


pulse_df |> 
  ggplot(aes(x = visit, y = bdi)) +
  geom_boxplot()

pulse_df |> 
  group_by(visit) |> 
  summarise(
    mean_bdi = mean(bdi, na.rm =TRUE),
    median_bdi = median(bdi, na.rm = TRUE)
  ) |> 
  knitr::kable(digits=2)

```








