---
title: "viz_2"
output: html_document
date: "2025-09-30"
---

```{r}
library(tidyverse)
library(patchwork)
library(p8105.datasets)
data("weather_df")
```

As a starting point, let’s revisit the scatterplot of tmax against tmin made in Visualization Pt 1.

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5)
```

## Labels 

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package"
  )
```

## Scales 

```{r}
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15"))

#Another way to do limits:

weather_df |> 
  filter(tmax > 10 , tmax < 30) |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15"))

######### Limits
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15ºC", "0", "15"),
    limits = c(-20, 30)) + 
  scale_y_continuous(
    trans = "sqrt", 
    position = "right")


#Changing the colors manually using scale_color_hue

weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15")) + 
  scale_color_hue(h=c(100,300))

#  scale_color_hue(h=c(100,300))



#Better way - Jeff suggests always using viridis
ggp_temperature = 
weather_df |> 
  ggplot(aes(x = tmin, y = tmax)) + 
  geom_point(aes(color = name), alpha = .5) + 
  labs(
    title = "Temperature plot",
    x = "Minimum daily temperature (C)",
    y = "Maxiumum daily temperature (C)",
    color = "Location",
    caption = "Data from the rnoaa package") + 
  scale_x_continuous(
    breaks = c(-15, 0, 15), 
    labels = c("-15º C", "0", "15")) + 
  viridis::scale_color_viridis(
    name = "Location", 
    discrete = TRUE
  )


ggsave("weather_scatterplot.png", ggp_temperature)


#saving it as ggp_temperature so that you dont have to keep adding to this code specifically
#use this structure name = .... if you want to save specific graphics or if your code chunk is getting large 
```



Update my base plot

```{r}

# Jeff's classic and go to 
ggp_temperature + 
  theme_minimal() +
  theme(legend.position = "bottom")

# Another possibility 
ggp_temperature + 
  theme_dark() +
  theme(legend.position = "bottom")

# And another! 
ggp_temperature + 
  theme_classic() + 
  theme(legend.position = "bottom")


#they have packages where you can make it look like excel 2003 or wes anderson movies.

```

# This is big! 

## Add this code at the begining of every RMD file

### Setting Options 

Jeff puts this in every to automatically set how he wants his figures to look 

```{r}
library(tidyverse)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Adding data in geoms

```{r}
central_park_df = 
  weather_df |> 
  filter(name == "CentralPark_NY")


molokai_df =
  weather_df |> 
  filter(name == "Molokai_HI")


ggplot(data = molokai_df, aes(x = date, y = tmax, color = name)) + 
  geom_point()+
  geom_line(data = central_park_df)
```


## `patchwork`

Make three plots separatemes and combine using patchwork

```{r}
ggp_tmax_tmin = 
  weather_df |> 
  ggplot(aes(x = tmin, y = tmax, colour = name)) +
  geom_point(alpha = 0.5) +
  theme(legend.position = "none")


ggp_prcp_density = 
  weather_df |> 
  filter(prcp > 0) |>                   #because there is that huge spike at 0. that we dont really care about 
  ggplot(aes(x = prcp, fill = name)) + 
  geom_density(alpha = .5) +
  theme(legend.position = "none")

ggp_temp_season =                       #shows seasonality
  weather_df |> 
  ggplot(aes(x = date, y = tmax, color = name)) + 
  geom_point(alpha = .5) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") 

(ggp_tmax_tmin + ggp_prcp_density)                    # this is the frame work for patchwork: this plus this place next to one another


(ggp_tmax_tmin + ggp_prcp_density) / ggp_temp_season  #you can divide to see them stacked on top of one another

```


## Data Manipulation 

Let's make temperature violin plots. 

```{r}
weather_df |> 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = 0.5)


# factor variables - 
```

GGplot will assign number to strings based on alphabetical order. In order to change this, use the following `mutate([] =fct_relevel([]))`


```{r}
weather_df |> 
  mutate(name = fct_relevel(name, c("Molokai_HI", "CentralPark_NY", "Waterhole_WA"))) |> 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = 0.5)

#fct_relevel says take your name variable and put the values w/i in the following order (L->R) reads (1->3) mutate this so this becuase the new [name] variable



# Another way that Jeff does it (and more frequently)
weather_df |> 
  mutate(name = fct_reorder(name, tmax)) |> 
  ggplot(aes(x = name, y = tmax, fill = name)) +
  geom_violin(alpha = 0.5)

#fact_reorder especially helpful if you have a ton of categories 
#here, organize [name] by ascending tmax value
```

#Tidy the data pre ggploting 
What about data tidiness?

Suppose I want to ask does the distribution of BDI indiex change across visits? 
To do this we need a box plot

BUT -- right now, or data are not structured this

```{r}
pulse_df =
  haven::read_sas("data/public_pulse_data.sas7bdat") |>  
  janitor::clean_names() |> 
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit",
    names_prefix = "bdi_score_",
    values_to = "bdi"
  ) |> 
  mutate(visit = replace(visit, visit == "bl", "00m"),
         visit = fct_inorder(visit)
        )  


#need to include haven:: becuase we didnt use library(haven) at begining 
# fct_inorder() saying order based on how it shows up in the dataset
```

# In class exercise 

How would I make the following graphic (look at the screen):

patchwork 
two different litters 7 on top and 8 on bottom 
dose (con, low, mod) on x axis
pn_day y axis (day on which this happened, post natal days for.....)

Looking at four different things: ears, pivot, eyes, walk

faceting for these two:  the groups (days of treatment) + outcome

## Make a plot for the FAS study
```{r}
pups_df = 
  read_csv("data/FAS_pups.csv", na = c("NA", ".", ""), skip = 3) |> 
  janitor::clean_names() |> 
  mutate(
    sex = case_match(
      sex,
      1 ~ "male",
      2 ~ "female"
    )
  )

litters_df = 
  read_csv("data/FAS_litters.csv", na = c("NA", ".", "")) |> 
  janitor::clean_names() |> 
  separate(group, into = c("dose","tx_day"), sep = 3)


fas_df = 
  left_join(pups_df, litters_df, by = "litter_number")

fas_df |> 
  select(pd_ears:tx_day) |> 
  pivot_longer(
    pd_ears:pd_walk,
    names_to = "outcome",
    names_prefix = "pd_",
    values_to = "pn_day"
  ) |> 
  mutate(outcome = fct_reorder(outcome, pn_day)) |> 
  drop_na() |> 
  ggplot(aes(x = dose, y = pn_day))+
  geom_violin() +
  facet_grid(tx_day ~ outcome)


#Note that the ggplot code is only three lines long. Everything else is data manipulation! 
```





